# Week 6 data analysis

This week we are analyzing to different datasets. Before going to the datasets. lets load a bunch of libraries beforehand to assist us.

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
```

Loads of the visualization etc. is reverse of the MABS material.

## The RATS data
### Loading the dataset and brief explorations

There appears to be some issues with the loading and dealing with the data (I couldn't get this to work no matter which way I tried), so this will go through some points of the wrangling again to produce the long data for RATS.

For this part we will implement *MABS chapter 8* analysis on the data.

```{r}
RATS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", header = TRUE, sep = '\t')
RATS$ID <- factor(RATS$ID)
RATS$Group <- factor(RATS$Group)
RATSL <- RATS %>% gather(key = WD, value = rats, -ID, -Group) %>% mutate(Time = as.integer(substr(WD,3,4))) 
```

To get us some general numerical information about the dataset, lets examine the data briefly as we've done in past weeks.

```{r}
str(RATSL)
head(RATSL)
glimpse(RATSL)
```

At this point there is not much to interpret - we are dealing with the wrangled data that is in LONG form instead of wide. Lets proceed to the actual analysis

### Chapter 8 data plots + analysis
#### Basic plotting..

Lets plot the data next. From the following we can differentiate the seperate groups.

```{r}
ggplot(RATSL, aes(x = Time, y = rats, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$rats), max(RATSL$rats)))
```

In all groups there is growth as the time passes. Group 1 seems to be most steady over time though, with more growth in groups 2 and 3. In group 2 there is one rather clear outlier (the gigantic rat..). Also, group 3 has one individual with somewhat less size. Overall the groups seem to be different in general size (group 1 having especially small speciments.)


#### Standardization and new visualization

Next we go on to standardize the data, have a peek at the numeric data and then redo the visualization on the standardized data. To standrardize..

```{r}
RATSL <- RATSL %>%
  group_by(Time) %>%
  mutate(stdrats = (rats - mean(rats))/sd(rats) ) %>%
  ungroup()

```
And have a glimpse of the standardized data..

```{r}
glimpse(RATSL)
```

And to revisualize..

```{r}
ggplot(RATSL, aes(x = Time, y = stdrats, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  scale_y_continuous(name = "standardized rats")
```

As we look at new graphsthe usual effect of scaling. Personally, this chart is harder for me to interpret, but we can still see the diferences between groups and outliers as in the previous chart.

#### Summary graph

```{r}
n <- RATSL$week %>% unique() %>% length()

RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise( mean = mean(rats), se = sd(rats)/sqrt(n) ) %>%
  ungroup()

glimpse(RATSS)
```

```{r}
ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3 )) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.8,0.5)) +
  scale_y_continuous(name = "mean(rats) +/- se(rats)")
```

The DataCamp exercise states that "Note the considerable overlap in the mean profiles of the two treatment groups suggesting there might be little difference between the two groups in respect to the mean [dataset] values". This doesn't exactly happen in these (despite here being one more group). 

#### Outliers

We already spotted some outliers, now we extreminate them from the data.

```{r}
RATSL8S <- RATSL %>%
  filter(Time > 0) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(rats) ) %>%
  ungroup()
```

```{r}
glimpse(RATSL8S)
```

Next we will draw a boxplot.

```{r}
ggplot(RATSL8S, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(rats), WD 8-64")
```

The boxplot reveals thee for us. Then we filter out the revealed 3 outliers using values 250, 550 and 500 as filter points.

```{r}
RATSL8S1 <- filter(RATSL8S, (Group == 1 & mean > 250) | (Group == 2 & mean < 550 | Group == 3 & mean > 500))
```

Then we redraw the plot with outliers gone.

```{r}
ggplot(RATSL8S1, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(rats without outliers), WD 1-64")
```

From the newly drawn plots we get what we all propably expected - thinner boxes as the outliers are gone.

#### Anova code

And now we proceed to perform anova to the data.

```{r}
RATSL8S2 <- RATSL8S %>%
  mutate(baseline = RATS$WD1)
```

```{r}
fit <- lm(mean ~ baseline + Group, data = RATSL8S2)
```

```{r}
anova(fit)
```

From the results we can se "baseline" having statistical significance, but the group variable not, having value of 0.076.

## The BPRS data
### Loading the dataset and brief explorations

As with the RATS data, we will load it, redo some parts of the wrangling process and display some numeric data to get get a better idea before we start with plots and analysis.

As opposed to the RATS data, this one is now based on chapter 9 of MABS material.

```{r}
BPRS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/BPRS.txt", sep  =" ", header = T)
BPRS$treatment <- factor(BPRS$treatment)
BPRS$subject <- factor(BPRS$subject)
BPRSL <-  BPRS %>% gather(key = weeks, value = bprs, -treatment, -subject)
BPRSL <-  BPRSL %>% mutate(week = as.integer(substr(weeks,5,5)))
```

To get us some general numerical information about the dataset, lets examine the data briefly as we've done in past weeks.

```{r}
str(BPRSL)
head(BPRSL)
glimpse(BPRSL)
```

As with the RATS data, we are dealing with long data again, not wide data.

Just to clarify what we are dealing with here, BPRS is a set of symptom constructors that are used to evaluate if a patient might have schizophrenia.

### Chapter 9 data plots + analysis
#### Basic plotting..

As the DataCamp results in rather unreadable chart (at least I can't make any sense of it), we use the same plot method as with RATS. The treatement groups are neatly in diffrent charts now. This is much more practical as we are have two treatment sets to deal with.

```{r}
ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))
```

Interpreting the treatment groups visually is slightly challenging for me. There seems to be one clear outlier in the group 2 at least. However, as to the question if patients score goes up or down during the treatment is rather unclear. It seems to fluctuate.. Some have general decline whereas some go down for a while, just to go back up again. The data is rather interesting, but I have no expertise to give any substantial comments on it.

#### Linear model

Next will use again a multiple linear regression model, which we have done in the past weeks, on the data, with the familiar lm() command. Then we will have a look at the summary for the model.

```{r}
BPRS_reg <- lm(bprs ~ week + treatment, data = BPRSL)
summary(BPRS_reg)
```

The basic model we applied shows statistical significance based for week with *** stars. Treatment2 recieving a value of 0.661 only.

#### Random intercept model

The next step in the process is using a random intercept modle. For that, we need new library. We load the library, create the random intercept model and print out a summary of it. The difference here is that random intercept doesn't assume independence between repeated measurements in score.

```{r}
library(lme4)
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)
summary(BPRS_ref)
```

#### Random intercept + Random slope model

And the upcoming step is to create a random intercept and random slope model, check a summary, and perform ANOVA on both of them. 

```{r}
BPRS_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = FALSE)
summary(BPRS_ref1)
anova(BPRS_ref1, BPRS_ref)
```

The reason why we did this is that fitting the random intercept and slope models lets linear regression for for each individual. What this offers us is a way to observe the scores and the time. Lets proceed to the next part.

#### Random intercept and random slope model with interaction


```{r}
BPRS_ref2 <- lmer(bprs ~ week * treatment + (week | subject), data = BPRSL, REML = FALSE)
summary(BPRS_ref2)
anova(BPRS_ref2, BPRS_ref1)
```

Here is the plot for observed values..

```{r}
ggplot(BPRSL, aes(x = week, y = bprs, group = subject)) +
  geom_line() +
  scale_x_continuous(name = "Time in weeks") +
  scale_y_continuous(name = "bprs") +
  theme(legend.position = "top")
```

As that looks rather confusing (at least to me), we use the earlier plotting method.

```{r}
ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))
```

Next we will create a vector of fitted values.

```{r}
Fitted <- fitted(BPRS_ref2)

BPRSL <- BPRSL %>%
  mutate(Fitted)
```

And, finally, the plot with fitted values for BPRSL.

```{r}
ggplot(BPRSL, aes(x = week, y = Fitted, group = subject)) +
  geom_line() +
  scale_x_continuous(name = "Time in weeks") +
  scale_y_continuous(name = "bprs") +
  theme(legend.position = "top")
```

And with the one that makes more sense to me personally.

```{r}
ggplot(BPRSL, aes(x = week, y = Fitted, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))
```